Natural selection in the Health and Retirement Study

### 2023-09-15 hrs-selection.qmd
Add agefb regs and cohorts bar plot

@@

  rand <- rand |>
    mutate(.by = rabyear,
      mean_raevbrn = mean(raevbrn, na.rm = TRUE),
      rlrs = raevbrn/mean_raevbrn
    ) |>
    mutate(
      income_resid = calc_income(rand),
      income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
      income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
      # as.numeric avoids haven.labelled issues:
      raedyrs = as.numeric(raedyrs),
      edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
      # wave 10 is 2010
      # table(rands$white$r10mstat)
       # 1                 1.married
       # 2   2.married,spouse absent
       # 3               3.partnered
       # 4               4.separated
       # 5                5.divorced
       # 6      6.separated/divorced
       # 7                 7.widowed
       # 8           8.never married
      married = factor(r10mstat == 1, 
                       levels = c(FALSE, TRUE),
                       labels = c("Other", "Married")),
    )

  rand <- rand |> left_join(rand_fam, 
                            by = join_by(hhid, pn),
                            unmatched = "drop",
                            relationship = "one-to-one")
  rand <- rand |> mutate(
    r10sibs = r10livsib + r10sbdied
  )

@@

tidy_pgs <- mod_pgs <- list()
tidy_pgs_ed <- mod_pgs_edyrs <- list()
tidy_pgs_sibs <- mod_pgs_sibs <- list()
tidy_pgs_x_ed <- mod_pgs_x_ed <- list()
tidy_pgs_x_inc <- mod_pgs_x_inc <- list()
tidy_pgs_x_mar <- mod_pgs_x_mar <- list()
cors_fert_edu <- tidy_pgs_joined <- list()
cors_sibs_edu <- list()
for (eth in eths) {
  rand <- rands[[eth]]

@@

  tidy_pgs_x_mar[[eth]] <-
    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_mar[[eth]]) <- pgs
  tidy_pgs_x_mar[[eth]] <-
    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
    filter(! term %in% c("(Intercept)", "marriedMarried", pcs)) |>
    mutate(
      married = str_remove(term, ".*:married")
    ) 
}

@@

titles <- list(white = "Whites (non-Hispanic)", black = "Blacks")
for (eth in eths) {
  ggp <- ggplot(tidy_pgs_joined[[eth]], aes(estimate.edyrs, estimate.fert)) +
          geom_hline(yintercept = 0, color = "grey80") +
          geom_vline(xintercept = 0, color = "grey80") +
          geom_point() +
          geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
                      linetype = "dashed", colour = "red", linewidth = 0.5) +
          scale_y_continuous(breaks = seq(-0.35, 0.15, 0.05),
                             labels = scales::label_number(0.01)) +
          coord_cartesian(xlim = c(-.3, .75)) +
          theme_light() + 
          theme(panel.grid = element_blank()) +
          labs(
            title = titles[[eth]],
            x = "Effect on years of education",
            y = "Effect on realized fertility"
          )
  print(ggp)
}

@@

@fig-x-edu shows that for both whites and blacks, there is essentially no difference in the distribution of absolute effect sizes between respondents with more and less education. @fig-x-inc, on the other hand, shows that absolute effect sizes appear larger among lower-income respondents, for both whites and blacks. However, in bootstrap tests, the median difference in absolute effect sizes is not significantly different from 0, for either ethnic group. @fig-x-mar shows absolute effect sizes for respondents who were married in 2010 against all other statuses. Among blacks, effect sizes look similar across marital status, but among whites they are larger for unmarried respondents, and this is significant in bootstraps (95% C.I. for median difference `r ci_mar_w[1]` to `r ci_mar_w[2]`).

@@

tidy_pgs_x_ed |> 
  list_rbind(names_to = "ethnicity") |> 
  mutate(
    Ethnicity = unlist(titles[ethnicity]),
    Education = str_remove(edu, "edu")
  ) |> 
  ggplot(aes(Education, abs(estimate), color = Education, fill = Education)) +
    geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
    geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
    facet_wrap(vars(Ethnicity), scales = "free_y") +
    labs(
      x = "Years of education",
      y = "Absolute effect size"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none"
    )
```

@@

@@

>>>>

titles <- list(white = "Whites (non-Hispanic)", black = "Blacks")

@@

  rand <- rand |> left_join(rand_fam, 
                            by = join_by(hhid, pn),
                            unmatched = "drop",
                            relationship = "one-to-one")
  rand <- rand |>
    mutate(.by = rabyear,
      mean_raevbrn = mean(raevbrn, na.rm = TRUE),
      rlrs = raevbrn/mean_raevbrn
    ) |>
    mutate(
      income_resid = calc_income(rand),
      income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
      income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
      # as.numeric avoids haven.labelled issues:
      raedyrs = as.numeric(raedyrs),
      edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
      # wave 10 is 2010
      # table(rands$white$r10mstat)
       # 1                 1.married
       # 2   2.married,spouse absent
       # 3               3.partnered
       # 4               4.separated
       # 5                5.divorced
       # 6      6.separated/divorced
       # 7                 7.widowed
       # 8           8.never married
      married = factor(r10mstat == 1, 
                       levels = c(FALSE, TRUE),
                       labels = c("Other", "Married")),
    r10sibs = r10livsib + r10sbdied,
    agefb = r10agey_e - h10ageokid,
    agefb = ifelse(agefb < 12, NA_real_, agefb),
    agefb = chop_equally(agefb, 2, labels = c("Low", "High"))
  )

@@

tidy_pgs <- mod_pgs <- list()
tidy_pgs_ed <- mod_pgs_edyrs <- list()
tidy_pgs_sibs <- mod_pgs_sibs <- list()
tidy_pgs_x_ed <- mod_pgs_x_ed <- list()
tidy_pgs_x_inc <- mod_pgs_x_inc <- list()
tidy_pgs_x_mar <- mod_pgs_x_mar <- list()
tidy_pgs_x_agefb <- mod_pgs_x_agefb <- list()

cors_fert_edu <- tidy_pgs_joined <- list()
cors_sibs_edu <- list()
for (eth in eths) {
  rand <- rands[[eth]]

@@

  tidy_pgs_x_mar[[eth]] <-
    map(mod_pgs_x_mar[[eth]], broom::tidy, conf.int = TRUE)
  names(tidy_pgs_x_mar[[eth]]) <- pgs
  tidy_pgs_x_mar[[eth]] <-
    list_rbind(tidy_pgs_x_mar[[eth]], names_to = "pgs") |>
    filter(! term %in% c("(Intercept)", "marriedMarried", pcs)) |>
    mutate(
      married = str_remove(term, ".*:married")
    )

  form_pgs_x_agefb <- paste0(pgs, ":agefb") |>
    map(\(x) reformulate(c(x, "agefb", pcs), response = "raevbrn"))
  mod_pgs_x_agefb[[eth]]  <-
    map(form_pgs_x_agefb, 
        \(x) lm(x, data = rands[[eth]], weights = weight)
        )

  tidy_pgs_x_agefb[[eth]] <-
    map(mod_pgs_x_agefb[[eth]], broom::tidy, conf.int = TRUE) |> 
    setNames(pgs) |> 
    list_rbind(names_to = "pgs") |>
    filter(! term %in% c("(Intercept)", "agefbHigh", pcs)) |>
    mutate(
      agefb = str_remove(term, ".*:agefb")
    )

}

@@

for (eth in eths) {
  ggp <- ggplot(tidy_pgs_joined[[eth]], aes(estimate.edyrs, estimate.fert)) +
          geom_hline(yintercept = 0, color = "grey80") +
          geom_vline(xintercept = 0, color = "grey80") +
          geom_point() +
          geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
                      linetype = "dashed", colour = "red", linewidth = 0.5) +
          scale_y_continuous(breaks = seq(-0.35, 0.15, 0.05),
                             labels = scales::label_number(0.01)) +
          coord_cartesian(xlim = c(-.3, .75)) +
          theme_light() + 
          theme(panel.grid = element_blank()) +
          labs(
            title = titles[[eth]],
            x = "Effect on years of education",
            y = "Effect on realized fertility"
          )
  print(ggp)
}

@@

@fig-x-edu shows that for both whites and blacks, there is essentially no difference in the distribution of absolute effect sizes between respondents with more and less education. @fig-x-inc, on the other hand, shows that absolute effect sizes appear larger among lower-income respondents, for both whites and blacks. However, in bootstrap tests, the median difference in absolute effect sizes is not significantly different from 0, for either ethnic group. @fig-x-mar shows absolute effect sizes for respondents who were married in 2010 against all other statuses. Among blacks, effect sizes look similar across marital status, but among whites they are larger for unmarried respondents, and this is significant in bootstraps (95% C.I. for median difference `r ci_mar_w[1]` to `r ci_mar_w[2]`). @fig-x-agefb splits respondents by age at first birth (median within each ethnicity). There are no large differences between early and late parents. Note that the N is lower here due to missing values. TODO test

@@

nobs <- map_dbl(eths, 
                \(x) nobs(mod_pgs_x_agefb[[x]][[1]])) |> 
  setNames(eths)

tidy_pgs_x_ed |> 
  list_rbind(names_to = "ethnicity") |> 
  mutate(
    Ethnicity = paste0(unlist(titles[ethnicity]), 
                       " (N = ", nobs[ethnicity], ")"),
    Education = str_remove(edu, "edu")
  ) |> 
  ggplot(aes(Education, abs(estimate), color = Education, fill = Education)) +
    geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
    geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
    facet_wrap(vars(Ethnicity), scales = "free_y") +
    labs(
      x = "Years of education",
      y = "Absolute effect size"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none"
    )
```

@@

```{r}
#| label: fig-x-agefb
#| fig-cap: |
#|   Violin plots of polygenic score effects on fertility by age at 
#|   first birth (median split). Effect sizes are
#|   absolute sizes of effects of a one standard deviation (within-ethnicity) 
#|   change of the PGS on fertility, controlling for 10 principal 
#|   components of genetic array data, estimated within the low/high income
#|   group. Lines show quartiles.

tidy_pgs_x_agefb |> 
    list_rbind(names_to = "ethnicity") |> 
    mutate(
        Ethnicity = unlist(titles[ethnicity]),
        agefb = fct_relevel(agefb, "Low", "High")
    ) |> 
    ggplot(aes(agefb, abs(estimate), color = agefb, fill = agefb)) +
      geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
      geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
      facet_wrap(vars(Ethnicity), scales = "free_y") +
      labs(
          x = "Age at first birth",
          y = "Absolute effect size"
      ) +
      theme_minimal() +
      theme(
          legend.position = "none"
      )

@@

# Appendix

```{r}
#| label: fig-cohorts
#| fig-cap: Distribution of birth years for the sample
#| fig-subcap:
#|   - ""
#|   - ""
#| layout-nrow: 2
#| column: page

for (eth in eths) {
  ggp <- ggplot(rands[[eth]], aes(rabyear)) + 
    geom_bar() +
    theme_minimal() +
    labs(x = "Birth year", y = "Count", title = titles[[eth]])
  print(ggp)
}
```
